---
title: "Lipinski rule over environmental chemical space"
author: "David Ross Hall"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
  fig_caption: true
    number_sections: false
    number_sections: true
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  error = FALSE
)
library(webchem)
library(tidyverse)
```

### David Hall --- created on 2022-09-02

# Lipinski's rule of 5

*As outlined on [wikipedia](https://en.wikipedia.org/wiki/Lipinski%27s_rule_of_five)*

 - No more than 5 hydrogen bond donors (the total number of nitrogen–hydrogen and oxygen–hydrogen bonds)
 - No more than 10 hydrogen bond acceptors (all nitrogen or oxygen atoms)
 - A molecular mass less than 500 daltons
 -An octanol-water partition coefficient (log P) that does not exceed 5



# Importing datasets

- Ripped EPA's Toxic Substances Control Act (TSCA) Chemical Substance Inveoty from steven. 
- Ditto w/ the inventory of ToxCast compounds
- drugs from [here](https://www.cureffi.org/2013/10/04/list-of-fda-approved-drugs-and-cns-drugs-with-smiles/)

```{r import}

tsca_raw <- read_csv("data/TSCAINV_202202.csv") %>%
  janitor::clean_names()

toxcast <- read_csv("data/toxcast_inv.csv") %>%
  janitor::clean_names()

drugs_raw <- read_tsv("data/drugs.tsv") %>%
  janitor::clean_names()

```

# Expanding TSCA database


- Removed *Unknown or Variable Compositions, Complex reaction products, and Biological materials)* (UVCBs) from TSCA. Defintions can be found [here](https://www.epa.gov/sites/default/files/2015-05/documents/uvcb.tsca <- tsca_raw %>%filter(is.na(uvcb))pdf)
- `flag` refers to compounds flagged in the TSCA. The meaning of EPA regulatory flags can be found [here](https://www.epa.gov/tsca-inventory/how-access-tsca-inventory#flags).
- `activity` refers to the commercial activity status.



```{r}
# getting logp values and cas from TSCA

tsca <- tsca_raw %>%
  filter(is.na(uvcb))

tsca_test <- sample_n(tsca, 500)

tsca_cid <- get_cid(tsca_test$casrn) %>%
  filter(!is.na(cid)) 

tsca_props <- pc_prop(cid = tsca_cid$cid,
                      properties = c("XLogP", 
                                     "MolecularFormula", 
                                     "MolecularWeight",
                                     "HBondDonorCount",
                                     "HBondAcceptorCount"))

# Function to add column indicating if molecule obeys Lipinski's rules
is_lipinski <- function(df, h_donor, h_accept, mw, logp){
  
  df <- df %>%
    mutate(lipinski = case_when(.data[[h_donor]] <= 5 & .data[[h_accept]] <= 10 & .data[[mw]] <= 500 & .data[[logp]] <= 5 ~ TRUE, TRUE ~ FALSE))
     
  df
    
}

tsca_lipinski <- tsca_props %>%
  is_lipinski(df = .,
              h_donor = "HBondDonorCount",
              h_accept = "HBondAcceptorCount",
              mw = "MolecularWeight",
              logp = "XLogP")


# combining all datasets together
tsca_all <- tsca_lipinski %>%
  mutate(CID = as.character(CID)) %>%
  left_join(., tsca_cid, by = c("CID" = "cid")) %>%
  rename(casrn = query) %>%
  left_join(., tsca_test, by = c("casrn" = "casrn")) %>%
  janitor::clean_names() %>%
  mutate(molecular_weight = as.numeric(molecular_weight))

  
```

```{r}

p <- ggplot(data = tsca_all, 
       aes(x = molecular_weight,
           y = x_log_p,
           colour = lipinski)) +
  geom_point(alpha = 0.3) 

p1 <- ggExtra::ggMarginal(p, 
                          type = "density"
                          )
p1
```

